Astrophysical calibration of gravitational-wave detectors

### 2015-08-05 docs/cosmic_cal.tex
Some large chunks of text removed

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see \cite{Vitale:2012} and references therein. For \ac{aLIGO} and \ac{AdV},
at the frequency range used in this project, the error in the calibration using
the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

% Describe our model for the calibration
\MP{The gravitational wave strain is measured through the differential arm length, $\Delta L$, 
changes of the interferometer via $h = \Delta L(f) / L$, where $L$ is the full arm length.
Calibration is required to relate the actual measured interferometer error signal output
$e(f)$ to $\Delta L$. This relation is known as the length response function, $R(f,t)$,
defined such that
\begin{equation}
\Delta L(f) = R(f,t) e(f),
\end{equation}
where $R$ varies slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to
uncertainties. In this study we will assume an estimate of $R$ is available (although
in theory we could take on the role of estimating $R$ itself), but that it differs from the
truth through some unknown scale factor, $s$, so that
\begin{equation}
sh(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $s > 1$ then a signal would appear to have a larger amplitude (e.g.\ be closer) than 
expected, whereas if $s < 1$ would appear to have a lower amplitude than expected. In this analysis 
we will simplify the situation by assuming that $s$ is a constant with respect to frequency, but in 
future studies that assumption could be dropped and $s$ could take some functional form, or 
piece-wise fit, with respect to $f$.}

@@

\MP{We want to assess how well a calibration scale factor can be estimated
from a single observed \ac{GW} associated with a particular \ac{sGRB}. As we {\it a priori}
have no knowledge of the likely location and distance of such an event we have performed
simulations of multiple events to see the expected distribution the accuracy of the
calibration scale factor recovery.}

% Describe the signal model (Taylor F2 etc...)
We use the TaylorF2 waveform approximation (see e.g.\ \cite{2009PhRvD..80h4043B} and references 
therein) with a 3.5 post-Newtonian expansion in phase for modelling our signal (in both simulations 
and signal recovery).  For \ac{BNS} systems we use non-spinning waveforms under the assumption that 
spins will be negligible. For \ac{NSBH} system we use a spinning waveform, but in which only the 
black hole has non-negligible spin and that its rotational angular momentum is aligned with the 
systems orbital angular momentum.

For a non-spinning system the general form of the frequency-domain polarisation amplitudes are
%
\begin{eqnarray}\label{eq:signal}
  \gws_{+}(f) &\propto \frac{1+\cos^{2}\iota}{D_{L}}
\mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\
  \gws_{\times}(f) &\propto
 \frac{\cos^{2}\iota}{D_{L}}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, 
\phi_{\mathrm{c}})}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$, with 
$\eta=m_{1}m_{2}/M^{2}$ and $M=m_{1}+m_{2}$, $\iota$ is the inclination angle, and $\nu=(\pi 
Mf)^{1/3}$.  The \ac{GW} strain measured at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^D(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+},F_{\times}$ are the antenna response functions which are dependent
upon the polarisation angle $\psi$ and the sky position of the source
$\alpha,\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we will consider advanced generation instruments operating
at their design sensitivity, \MP{with noise power spectral densities taken from
\cite{2013arXiv1304.0670L}}. The network we consider is the 3 detector H1--L1--V1 configuration.

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\tilde{n}_{k}(f)+s\gws_{k}(f,\btheta)
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of Gaussian noise drawn from a
distribution consistent with the power-spectral-density of the advanced
detector noise.

@@

\MP{Our aim is to estimate the marginal posterior probability distribution function (pdf)
for the unknown calibration scale factors. We use Bayes' theorem for which we need to define
a likelihood function and prior pdfs on the signal parameters being estimated. In this
analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to sample 
the posterior distribution.} We use a Gaussian likelihood function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
\fl p(\bi{d} | \btheta, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\Delta f$ is the frequency resolution of the data, $S_k$ is the $k^{\rm th}$ detector's one 
sided noise power spectral density, and the $i$ indices increment over frequency with $i_{f_{\rm 
low}}$ and $i_{f_{\rm high}}$ corresponding to the lower and upper range in frequencies used for 
our analysis, which were 20\,Hz to 400\,Hz respectively. This frequency range was chosen as the 
vast majority of the \ac{SNR} can be found within this range, so going to lower or higher 
frequencies provides very little additional information for our analysis. The source parameters, 
and the unknown scale factor, have been subsumed into a vector $\btheta$. For a network of 
detectors, the overall likelihood function becomes a product of the likelihoods for each detectors, 
i.e.\
\begin{equation}
  \label{eq:mult-likeli}
  p(\bi{d}| \btheta, \curlH) = \prod \limits_D p(\bi{d}^D| \vec{\theta}, \curlH)
\end{equation}
where $D = \{H,L,V\}$.

@@

The primary reason why we can use coincident observations with a GRB as a potential check of 
detector calibration is that the GRB allows us to constrain our priors on various parameters that 
would normally be highly correlated with the signal amplitude. The most important of these is that 
the GRB can provide a very tight constraint on the source distance\footnote{For \ac{BNS} and 
\ac{NSBH} systems they will be observable only in the relatively nearby Universe ($z \sim 
0.1$ and $z\sim 0.2$ for \ac{BNS} and \ac{NSBH} systems respectively), so the relation between the 
measured GRB redshift and distance is simple and our luminosity distance will be negligably 
effected by the choice of cosmology \MP{Check this?}.}. In this analysis we therefore assume that 
the source distance is known (i.e.\ has a $\delta$-function prior). Another piece of information 
that we can use from having a coincident GRB is that the system is likely to be relatively close to 
face-on (an therefore nearly circularly polarised), which allows us to place prior constraints on 
the system inclination angle. Beam opening angles for short GRBs are relatively poorly constrained 
as they are based on only a few jet break detections, however \cite{2014ApJ...780..118F} give a 
median opening angle value of $\sim 10^{\circ}$. If we assume that the beam opening angle is a 
half-normal distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma \sim 
0.67\sigma$, then a median of $\sim 10^{\circ}$ corresponds to $\sigma \sim 14.8^{\circ}$. We 
therefore place a Gaussian prior on the inclination angle with zero mean and $\sigma=14.8^{\circ}$.

Our prior on the system component masses is dependent on whether we are considering the source 
being a \ac{BNS} system or a \ac{NSBH}. For the former case the prior we use is a Gaussian 
distribution for both components with means of $1.35\,\textrm{M}_{\odot}$ and standard deviations 
of $0.13\,\textrm{M}_{\odot}$ \MP{(reference)}. In reality, for all the systems we use in our 
study, the chirp mass, which we see from equation~\ref{eq:signal} plays a role in the overall 
signal amplitude, is very well constrained by the systems phase evolution, so our prior could be 
expanded with minimal effect on the results. In the case of \ac{NSBH} systems we use the same 
prior as above for the neutron star, but for the black hole we use a prior based on the 
canonical masss distribution used for the rate results in \cite{2012PhRvD..85h2002A} with a mean of 
$5\,\textrm{M}_{\odot}$ and a standard deviation of $1\,\textrm{M}_{\odot}$. For the \ac{NSBH}
systems we also require a prior on the black hole spin (for \ac{BNS} systems we have assumed that 
the spins are small enough that they will be negligible). We use a uniform prior on the 
normalised aligned spin magnitude between $-1$ and $1$.

@@

For the spinning \ac{NSBH} systems we use a uniform prior on the dimensionless spin magnitude of 
the black hole between $-1$ and 1.

Finally, we require a prior on the calibration scale factors for each detector. We make the 
assumption that in general the calibration will be correct, so want to use a prior that is peaked 
at unity. We also assume that the probability density that the calibration scale factor is either 
larger or smaller than unity by an equivalent factor, e.g.\ $s=10$ or $s=0.1$, is the same. For this 
we therefore use a log-normal distribution as our prior
\begin{equation}
 p(s|I) = \frac{1}{s\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{s} - \mu)^2}{2\sigma^2} \right)},
\end{equation}
where we chose $\sigma = 1.07$, which in turn means that having the mode at unity gives $\mu = 
1.15$. With these parameters the prior probability density at 0.1 and 10 is one tenth of that at 
1. Note that this prior does not provide produce a symmetric amount of probability about unity, but 
given our simulation criterion described below in Section~\ref{sec:simulations} we find that the 
likelihood generally overwhelms the prior. As $s$ is a scale factor we also have the further 
constraint that it must be positive.

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 900\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations use the 
two Advanced LIGO detectors (one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced 
Virgo detector (V1). The simulations are all performed in the freqeuncy domain and span a frequency 
range from 20\,Hz to 400\,Hz. We assume all detectors are expected to be operating at their design
sensitivity (as given by Figure 1 of \cite{2013arXiv1304.0670L}). As such we add noise to each 
simulation based on the sensitivity curve, but scaled with the associated calibration scale factor.

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed up to 2000 simulations at each 
distance value\footnote{Simulations on which our MCMC did not produce converged chains within our 
analysis time were discarded and not re-run, although there were always at least 1000 simulations 
that completed.}. For proposing injections for each simulation the source parameters were randomly 
drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky position being 
drawn randomly from a uniform distribution, and with the scaling factors for each detector drawn 
from a Gaussian distribution with a mean of one and standard deviation of $0.125$ (equivalent to a 
mean fractional offset of 10\%). However, in accepting a proposed injection as one to be analysed 
we introduced a criteria that the signal be ``detectable'', based on it having an SNR of $\geq 5.5$ 
in at least two detectors as used in \cite{2012PhRvD..85h2002A} (as these would be signals 
coincident with a GRB we do not include the often used further constraint that the total coherent 
network SNR is greater than 12). The coalescence times for all injections were held fixed 
within the centre of the prior window. This criteria means that as we go to larger distances our 
population of injected sources will not be uniformly distributed over the sky, and will generally 
be closer to being circularly polarised. For the \ac{NSBH} systems it also means that at larger 
distances we will have more sources with larger mass ratios \MP{Check this}.

These simulations allow us to assess how well on average we would be able to constrain the 
calibration scale for a given CBC-GRB coincidence.

@@

A simple assumption that one could make would be that the uncertainties on the calibration scale 
factors (if they are independent of other parameters and have a roughly Gaussian probability 
distribtuion)
should be given by $\sim 1/{\rm SNR}$ for each detector. As we will show below this assumption is 
reasonable, but small correlations do exist between parameters meaning that it does not completely 
hold.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident GRB is observed and can yield a distance estimate). The boxplots span the lower to 
upper quartile range of the distributions, with the median value shown as a horizontal line within 
the box and the mean shown as a star. The dashed magenta line shows the percentage of source drawn 
from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{NSBH} systems 
(provided a coincident GRB is observed and can yield a distance estimate). The plot contents are 
the same as 
in figure~\ref{fig:bnsresults}.}
\end{figure}

In figure~\ref{fig:bnsresults} we see that unsurprisingly on average the scale factor can be 
recovered to equivalent precision in both \ac{aLIGO} detectors, with uncertainties generally within 
the 10\% range for sources at 100\,Mpc. An interesting feature that we see is that for distances 
$\gtrsim 250$\,Mpc the upper extent of the uncertainty hits a maximum at $\sim 25\mbox{--}30\%$, 
whilst the width of boxes narrows. This is due to our ``detectability'' criteria, in that for all 
distances we will only be seeing those sources with high enough SNR that we would consider them 
detectable. There will therefore be fewer sources with SNR higher than this criteria at large 
distances, giving us a narrower range, and we automatically exclude those with SNR that are too 
small thus truncating our uncertainty distribution at the upper range. However, this does show that 
on average for sources that are detectable out to 450\,Mpc we would be able to constrain the 
calibration scale factor uncertainty to $\sim 20\%$. As the detection criteria will generally come 
from the SNR being high enough within the two \ac{aLIGO} instruments this means that the SNR in 
\ac{AdV} can be small and thus the ability to constrain the scale factor for it becomes poor 
(although it still provides information that the calibration is not grossly inaccurate). We also 
see that true uncertainties achievable for recovering the calibration scale factors are indeed very 
similar to the simple assumption that they would be given by $sim 1/{\rm SNR}$. Although, they are 
probably slightly poorer due to their still be correlations with $\iota$ even within the 
constrained prior range, and slight correlation between the scale factors for H1 and L1 as seen in 
figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see very similar results although the higher \ac{SNR} of the 
signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on the calibration scale 
factors for H1 and L1 out to greater distances.

@@

This project is only the beginning in testing this new method of calibration.
More rigorous testing has to be carried out for a number of different systems.
As mentioned above, a larger number of random sky positions have to be analysed
to give better understanding of the resolution capabilities of the method. The
current technique of calibration defines the error in the detector to be a
complex function of frequency and phase of the GW. Instead of the simple scale
factor used so far, a more realistic error can be tested to ensure that the
estimation in its value is still comparable with current methods.

>>>>

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabret-Perot Michelson interferometer; where 
an elaborate feedback system is used to sustain a defined measurement in arm length difference 
between the moving mirrors. A comprehensive description of the calibration procedure (in particular 
for the LIGO detectors during their fifth science run) can be found in \cite{2010NIMPA.624..223A}, 
and a brief description is giving in \cite{Vitale:2012} and references therein. For \ac{aLIGO} 
and \ac{AdV}, at the frequency range used in this project, the error in the amplitude calibration 
using the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed new method in this project.~\cm{Clean this up and 
expand.}

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$s$, so that
\begin{equation}
sh(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $s > 1$ then a signal would appear to have a larger amplitude (e.g.\ be closer) than 
in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and the noise 
will be contain the scale factor}, whereas if $s < 1$ it would appear to have a lower amplitude 
than reality. In this analysis we will simplify the situation by assuming that $s$ is a constant 
with respect to frequency, but in future studies that assumption could be dropped and $s$ could 
take some functional form, or piece-wise fit, with respect to $f$.

@@

We want to assess how well the calibration scale factor can be estimated from a single observed 
\ac{GW} associated with a particular \ac{sGRB}. As we {\it a priori} have no knowledge of the 
likely location and distance of such an event we have performed simulations of multiple events to 
see the expected distribution the accuracy of the calibration scale factor recovery.

% Describe the signal model (Taylor F2 etc...)
We use the TaylorF2 waveform approximation (see e.g.\ \cite{2009PhRvD..80h4043B} and references 
therein) with a 3.5 post-Newtonian expansion in phase for modelling our signal (in both simulations 
and signal recovery).  For \ac{BNS} systems we use non-spinning waveforms under the assumption that 
spins will be negligible. For \ac{NSBH} system we use a spinning waveform, but in which only the 
black hole has non-negligible spin and that its rotational angular momentum is aligned with the 
system's orbital angular momentum.

For a non-spinning system the general form of the frequency-domain polarisation amplitudes are
%
\begin{eqnarray}\label{eq:signal}
  \gws_{+}(f) &\propto \frac{1+\cos^{2}\iota}{D_{L}}
\mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\
  \gws_{\times}(f) &\propto
 \frac{\cos^{2}\iota}{D_{L}}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, 
\phi_{\mathrm{c}})}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$, with 
$\eta=m_{1}m_{2}/M^{2}$ and $M=m_{1}+m_{2}$, $\iota$ is the inclination angle, and $\nu=(\pi 
Mf)^{1/3}$.  The \ac{GW} strain measured at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^k(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+},F_{\times}$ are the antenna response functions which are dependent upon the 
polarisation angle $\psi$ and the sky position of the source $\alpha$ and $\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we will consider the advanced generation LIGO and Virgo detectors operating
at their design sensitivity, with noise power spectral densities taken from
\cite{2013arXiv1304.0670L}. The network we consider is the two \ac{aLIGO} detectors (H1 and L1) and 
the \ac{AdV} detector (V1).

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\tilde{n}_{k}(f)+s\gws_{k}(f,\btheta)
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of Gaussian noise drawn from a distribution 
consistent with the \ac{PSD} of the advanced detector noise.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf) given the data contains a signal. We 
use Bayes' theorem for which we need to define a likelihood function and prior pdfs on the signal 
parameters being estimated. In this analysis we use the Markov chain Monte Carlo code {\tt emcee} 
\cite{2013PASP..125..306F} to sample the posterior distribution. We use a Gaussian likelihood 
function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
\fl p(\bi{d} | \btheta, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\bi{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's one sided noise \ac{PSD}, and the $i$ indices 
increment over frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters, and the unknown scale factors, have been subsumed 
into a vector $\btheta$.

@@

The primary reason why we can use coincident observations with a \ac{sGRB} as a potential check of 
detector calibration is that the \ac{GRB} allows us to constrain the priors on various parameters 
that would normally be highly correlated with the signal amplitude. The most important of these is 
that the \ac{GRB} can provide a very tight constraint on the source distance\footnote{For \ac{BNS} 
and \ac{NSBH} systems they will be observable only in the relatively nearby Universe ($z \sim 
0.1$ and $z\sim 0.2$ for \ac{BNS} and \ac{NSBH} systems respectively), so the relation between the 
measured GRB redshift and distance is simple and our luminosity distance will be negligibly 
effected by the choice of cosmology \MP{Check this?}.}. In this analysis we therefore assume that 
the source distance is known (i.e.\ has a $\delta$-function prior). Another piece of information 
that we can use from having a coincident \ac{GRB} is that the system is likely to be relatively 
close to face-on (an therefore nearly circularly polarised), which allows us to place prior 
constraints on the system inclination angle. Beam opening angles for \acp{sGRB} are relatively 
poorly constrained as they are based on only a few jet break detections, however 
\cite{2014ApJ...780..118F} give a median opening angle value of $\sim 10^{\circ}$. If we assume 
that the beam opening angle is a half-normal distribution, for which the median is given by 
$\sqrt{2}\erf{}^{-1}(1/2)\sigma \sim 0.67\sigma$, then a median of $\sim 10^{\circ}$ corresponds to 
$\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian prior on the inclination angle with zero 
mean and $\sigma=14.8^{\circ}$.

Our prior on the system component masses is dependent on whether we are considering the source 
being a \ac{BNS} system or a \ac{NSBH}. For the former case the prior we use is a Gaussian 
distribution for both components with means of $1.35\,\textrm{M}_{\odot}$ and standard deviations 
of $0.13\,\textrm{M}_{\odot}$ \MP{(reference)}. In reality, for all the systems we use in our 
study, the chirp mass, which we see from equation~\ref{eq:signal} plays a role in the overall 
signal amplitude, is very well constrained by the system's phase evolution, so our prior could be 
expanded with minimal effect on the results. In the case of \ac{NSBH} systems we use the same 
prior as above for the neutron star, but for the black hole we use a prior based on the 
canonical masss distribution used for the rate results in \cite{2012PhRvD..85h2002A} with a mean of 
$5\,\textrm{M}_{\odot}$ and a standard deviation of $1\,\textrm{M}_{\odot}$. For the \ac{NSBH}
systems we also require a prior on the black hole spin (for \ac{BNS} systems we have assumed that 
the spins are small enough that they will be negligible). We use a uniform prior on the 
normalised aligned spin magnitude between $-1$ and $1$.

@@

Finally, we require a prior on the calibration scale factors for each detector. We make the 
assumption that in general the calibration applied to the detector data will be correct, so want to 
use a prior that is peaked at unity. We also assume that the probability density that the 
calibration scale factor is either larger or smaller than unity by an equivalent factor, e.g.\ 
$s=10$ or $s=0.1$, is the same. For the scale factor for each detector we we therefore use a 
log-normal distribution as our prior
\begin{equation}
 p(s|I) = \frac{1}{s\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{s} - \mu)^2}{2\sigma^2} \right)},
\end{equation}
where we chose $\sigma = 1.07$, which in turn means that having the mode at unity gives $\mu = 
1.15$. With these parameters the prior probability density at 0.1 and 10 is one tenth of that at 
1. Note that this prior does not give a symmetric amount of probability about unity, but given our 
simulation criterion described below in Section~\ref{sec:simulations} we find that the likelihood 
generally overwhelms the prior. As $s$ is a scale factor we also have the further constraint that 
it must be positive.

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 900\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations used the 
two \ac{aLIGO} detectors (H1 and L1) and the \ac{AdV} detector (V1). The simulations were all 
performed in the frequency domain and spanned a frequency range from 20\,Hz to 400\,Hz. We assumed
all detectors were operating at their design sensitivity (as given by Figure~1 of 
\cite{2013arXiv1304.0670L}). As such we added noise to each simulation based on the sensitivity 
curve, but scaled with the associated calibration scale factor.

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed up to 2000 simulations at each 
distance value\footnote{Simulations on which our MCMC did not produce converged chains within our 
analysis time were discarded and not re-run, although there were always at least 1000 simulations 
that completed.}. For proposing injections for each simulation the source parameters were randomly 
drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky position being 
drawn randomly from a uniform distribution, the coalescence times held fixed within the centre of 
the prior window, and with the calibration scaling factors for each detector drawn from a Gaussian 
distribution with a mean of one and standard deviation of $0.125$ (equivalent to a mean fractional 
offset of 10\%). However, in accepting a proposed injection as one to be analysed we introduced a 
criteria that the signal be ``detectable'', based on that used in \cite{2012PhRvD..85h2002A} of it 
having an \ac{SNR} of $\geq 5.5$ in at least two detectors (as these would be signals coincident 
with a \ac{GRB} we do not include the often used further constraint that the total coherent network 
\ac{SNR} is greater than 12).  This criteria means that at larger distances our population of 
injected sources was not be uniformly distributed over the sky, and were generally closer to being 
circularly polarised.

These simulations have allowed us to assess how well on average we would be able to constrain the 
calibration scale for a given \ac{CBC}-\ac{GRB} coincidence.

@@

A simple assumption that one could make would be that the uncertainties on the calibration scale 
factors (if they are independent of other parameters and have a roughly Gaussian probability 
distribution) should be given by $\sim 1/{\rm SNR}$ for each detector. As we will show below this 
assumption is reasonable, but small correlations do exist between parameters meaning that it does 
not completely hold.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident \ac{GRB} is observed and can yield a distance estimate). The boxplots span the lower 
to upper quartile range of the distributions, with the median value shown as a horizontal line 
within the box and the mean shown as a star. The dashed magenta line shows the percentage of source 
drawn from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{NSBH} systems 
(provided a coincident \ac{GRB} is observed and can yield a distance estimate). The plot contents 
are the same as in figure~\ref{fig:bnsresults}.}
\end{figure}

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty hits a maximum at $\sim 
25\mbox{--}30\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still be correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH} systems although the 
higher \ac{SNR} of the signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on 
the calibration scale factors for H1 and L1 out to greater distances.

\MP{Here's my current thinking regarding why the distributions for the NSBH systems are broader
that those for the BNS systems. At distances where the distributions in SNR for each type of system 
are equivalent for the NSBH systems the ``detectable'' systems can be produced
by a broader range of parameter combinations, in particular chirp mass and $\iota$, this means
that the particular correlations between these are the scale factors will mean that as a whole
the distribution will be broader. I need to convince myself of this.}

@@
